Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-03 journal paper/content/sec-intro.tex
Some more edits.

%We investigate these questions with 
\xhdr{Competition game.} We consider a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the basic model is as follows:

@@

We distinguish three classes of bandit algorithms, based on 
the prevalent intuition in the area. The distinction concerns the way in which they resolve the fundamental tradeoff between  exploration and \emph{exploitation} (making optimal decisions using the available data). Going from more primitive to more sophisticated:

@@

\item \emph{Exploration-separating algorithms}
 that separate exploration and exploitation: each round is dedicated to one and completely ignores the other.

\item \emph{Adaptive exploration} algorithms that combine exploration and exploitation, and sway the exploration choices towards more promising alternatives.
\end{itemize}

\noindent We are mainly interested in qualitative differences between the three classes. For numerical simulations, we pick one representative algorithm from each class. For theoretical results, we allow arbitrary algorithms and focus on asymptotic differences in the algorithms' performance.

@@

We find that competition induces firms to commit to a greedy (myopic) algorithm in equilibrium that does no purposeful exploration. This leads to low consumer welfare since the greedy algorithm is known to be dramatically bad in many important cases of multi-armed bandits. The primary mechanism that generates this stark result is that consumers need to be incentivized to select a firm over its competitors, leading a firm that engages in exploration to be starved of consumers before it makes enough progress on its learning problem. In order to incentivize ``better" exploration strategies in equilibrium, the key intuition is that the firm needs to have some ``free" consumers that visit them without the firm having to incentivize them to do so. This allows the firm to eventually overcome the initial losses in consumer perception from exploration. We explore two economic mechanisms that can generate such an effect.

>>>>

%We investigate these questions with 
\xhdr{Competition game.} We consider a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the key features of our model are as follows:

@@

We distinguish three classes of bandit algorithms, based on 
the prevalent intuition in the area. The distinction concerns the way in which they resolve the fundamental tradeoff between  exploration and \emph{exploitation} (making optimal myopic decisions using the available data). Going from more primitive to more sophisticated, these three classes are as follows:

@@

\item \emph{Exploration-separating algorithms}
 that separate exploration and exploitation: essentially, each round is dedicated to one and completely ignores the other.

\item \emph{Adaptive-exploration} algorithms that combine exploration and exploitation, and gradually sway the exploration choices towards more promising alternatives.
\end{itemize}

\noindent Greedy algorithms are terrible for a wide variety of problem instances, precisely because they never explore. Adaptive-exploration algorithms are optimal in the worst case, and exponentially improve for ``easy" problem instances. Exploration-separated algorithms learn at a reasonable but mediocre rate across all problem instances, and fail to improve for the ``easy" ones.

@@

We offer a mix of theoretical results and numerical simulations. We are mainly interested in qualitative differences between the three classes of algorithms in Section~\ref{sec:intro-model}. For numerical simulations, we pick one representative algorithm from each class. For theoretical results, we allow arbitrary algorithms and focus on asymptotic differences in the algorithms' performance.

\xhdr{Main results.}
We start with a basic version in which the principals are symmetric (\ie do not have an inherent advantage over one another) and the agents are utility-maximizers. We find that competition induces firms to choose the greedy algorithm.

commit to a greedy (myopic) algorithm in equilibrium that does no purposeful exploration. This leads to low consumer welfare since the greedy algorithm is known to be dramatically bad in many important cases of multi-armed bandits. The primary mechanism that generates this stark result is that consumers need to be incentivized to select a firm over its competitors, leading a firm that engages in exploration to be starved of consumers before it makes enough progress on its learning problem. In order to incentivize ``better" exploration strategies in equilibrium, the key intuition is that the firm needs to have some ``free" consumers that visit them without the firm having to incentivize them to do so. This allows the firm to eventually overcome the initial losses in consumer perception from exploration. We explore two economic mechanisms that can generate such an effect.
